---
title: "Danish Kings"
author: "Manuel Ø. Mørch"
date: "2026-03-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```



## Load the kings


1. Look at the dataset that are you loading and check what its columns are separated by?


*In monarchs.csv the seperator is "," *

2. Create a `kings` object in R with the different functions below and inspect the different outputs.
-  `read.csv()`
-  `read_csv()`
-  `read.csv2()` 
-  `read_csv2()`
```{r eval=TRUE}

# FILL IN THE CODE BELOW and review the outputs
kings1 <- read.csv("data/monarchs.csv", na="NULL")
kings1
# kings 1 Works fine but it is not a tibble, it is a data frame.

kings2 <- read_csv("data/monarchs.csv", na="NULL")
kings2
# kings2 Works perfectly and it is a tibble.

kings3 <- read.csv2("data/monarchs.csv", na="NULL")
kings3
# kings3 does not work, it reads the whole dataset as one column

kings4 <- read_csv2( "data/monarchs.csv", na="NULL")
kings4
# kings4 does not work, it reads the whole dataset as one column
```
Answer: *kings2 is what works for my .csv file*


1. Which of these functions is a `tidyverse` function? Read data with it below into a `kings` object

Answer: *read_csv is a tidyverse function.*

2. What is the result of running `class()` on the `kings` object created with a tidyverse function.

Answer: *"spec_tbl_df" "tbl_df"      "tbl"         "data.frame" which tells us it is a tibble*

3. How many columns does the object have when created with these different functions?

Answer: *read.csv created 5 columns, read_csv created 5 columns, read.csv2 created 1 column and read_csv2 created 1 column.*


4. Show the dataset so that we can see how R interprets each column

*Where i wrote head(kings) you can see that it interprets all columns as characters, so i have to remember to coerce the start and end of reign columns to numbers before i can do calculations on them.*



```{r eval = TRUE}
# COMPLETE THE BLANKS BELOW WITH YOUR CODE, then turn the 'eval' flag in this chunk to TRUE.
library(tidyverse)
kings <- read_csv("data/monarchs.csv", na="NULL")

class(kings)

head(kings)


glimpse(kings)


```


## Calculate the duration of reign for all the kings in your table



Create a new column called  `duration` in the kings dataset, utilizing the `mutate()` function from tidyverse.

*(I just had to convert it so it understood my reign_start and reign_end as numbers, not characters, and then I could just subtract them to get the duration of reign in years.)*

```{r, eval=TRUE, warning=FALSE}
# Calculating duration

#first i have to make sure it reads the data correctly, so everything except for the names should be read as numbers, not characters.
kings <- kings %>%
  mutate(across(-name, as.numeric))
# Then i can calculate the duration of reign by subtracting the start from the end of reign.

kings <- kings %>%
  mutate(duration = reign_end - reign_start)
kings

```

## Calculate the average duration of reign for all rulers
So you need to subset your `kings` dataset to the `duration` column. If you subset it as a vector you can calculate average on it with `mean()` base-R function. If you subset it as a tibble, you can calculate average on it with `summarize()` tidyverse function. Try both ways!

* You first need to know how to select the relevant `duration` column. What are your options?
Answer: *You can select the duration column with either `kings$duration` or with `select(duration)`*
* Is your selected `duration` column a tibble or a vector? The `mean()` function can only be run on a vector. The `summarize()` function works on a tibble.
Answer: *If i select the duration column with `kings$duration` it is a vector, if i select it with `select(duration)` it is a tibble.*
* Are you getting an error that there are characters in your column? Coerce your data to numbers with `as.numeric()`.
*I already did that in the previous step*
* Remember to handle NAs: `mean(X, na.rm=TRUE)`
*done*

```{r eval=TRUE, warning=FALSE}
# Calculating average duration

# Vector approach
avg_years_ruled_vector <- mean(kings$duration, na.rm = TRUE)
avg_years_ruled_vector

# Tibble approach
avg_years_ruled_tibble <- kings %>%
  summarize(avg_years_ruled = mean(duration, na.rm = TRUE))
# Here i also have to use pull() so it will display just the number instead of a tibble
pull(avg_years_ruled_tibble)



```

*The average duration of reign is `r avg_years_ruled_tibble` years.*

## How many and which kings enjoyed a longer-than-average duration of reign?

Display the result and also  count the resulting rows with `count()`

```{r}
# longer-than-average duration of reign

# Because i have to use pull() to just get the number with the tibble version, it is best to just use the vector version here.
# I will also arrange them in descending order. Just to make it nice
kings %>% 
  filter(duration > avg_years_ruled_vector) %>%
  select(name, duration) %>%
  arrange(desc(duration))

# and to count how many had longer-than-average reign:
kings %>% 
  filter(duration > avg_years_ruled_vector) %>%
  count()

```


## How many days did the three longest-ruling monarchs rule? 

* Sort kings by reign `duration` in the descending order. Select the three longest-ruling monarchs with the `slice()` function
* Use `mutate()` to create `Days` column where you calculate the total number of days they ruled
* BONUS: consider the transition year (with 366 days) in your calculation!

```{r, eval=TRUE}
# YOUR CODE
# Monarchs sorted by reign duration in descending order. Then selecting the tree longest-ruling monarchs
kings %>% 
  arrange(desc(duration)) %>% 
  slice(1:3) %>% 
# Calculating the total number of days they ruled while considering the transition year (with 366 days) in the calculation.
#Transition years/leap years are roughly every 4 years, so i just use 365.25 instead of 365.
  mutate(days = duration * 365.25) %>% 
  select(name, duration, days)
```


## Challenge: Plot the kings' duration of reign through time

What is the long-term trend in the duration of reign among Danish monarchs? How does it relate to the historical violence trends ? 


```{r, warning=FALSE}
# YOUR CODE

mutate(kings, midyear = reign_start + ((reign_end - reign_start)/2)) %>% 
  ggplot(aes(x = midyear, y = duration)) +
  geom_point() +
  geom_smooth() +
  labs(title = "Duration of Reign of Danish Monarchs Over Time",
       x = "Midyear of Reign",
       y = "Duration of Reign (years)")
```


